---
title: "Lab 4"
subtitle: "ECOL 620 - Applications in Landscape Ecology"
author: "George Woolsey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    # code_folding: hide
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding){ 
    out_dir <- '../';
    rmarkdown::render(inputFile, encoding = encoding, output_file=file.path(dirname(inputFile), out_dir, 'index.html'));
    file.copy(from = "..//index.html", to = '..///data//lab4_george_woolsey.html', overwrite = TRUE)
  })
  
---

# Setup

```{r, include=FALSE, warning=F, message=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results='hide'
  , fig.width = 10
  , fig.height = 7
)
```

```{r, eval=T}
# bread-and-butter
library(tidyverse)
library(lubridate)
library(viridis)
library(scales)
library(latex2exp)
# visualization
library(ggrepel)
library(cowplot)
library(kableExtra)
# spatial analysis
library(USAboundaries)
library(sf)
library(terra)
library(landscapemetrics) #for patch, class, and landscape metrics
library(FedData) #for nlcd color pallette
# set seed
set.seed(11)
```

# Rules

1. Using the provided R Markdown template, answer all questions and show your R code where necessary. Note, some questions are just plain text written answers.
2. Complete your assignment using the R markdown file and submit individual assignments to Canvas. Knit your script and submit an .html file on Canvas. Please use  the following naming convention: lab4_firstname_lastname.html (ex. lab4_kyle_horton.html).

# Short Answer

## Question 1

Briefly describe (1-2 sentences) what each of the following functions achieve. All of these functions are seen throughout the Lab #3 example code. (1 pt each). 

* `calculate_lsm` is from the `landscapemetrics` package and can be used to calculate several landscape metrics at the user-defined level (i.e. patch, class, or landscape). Landscape metrics include patch area, number of patches, percentage of landscape, and radius of gyration, for example (see `landscapemetrics::list_lsm`).

* `lsm_c_np` is from the `landscapemetrics` package and can be used to calculate the total number of patches (an aggregate metric) at the *class* level.

* `show_cores` is from the `landscapemetrics` package and can be used to create a plot with the core area of patches where the core area is the area of the patch excluding edges of user-defined depth.

* `lsm_l_shei` is from the `landscapemetrics` package and can be used to calculate Shannon's evenness index (a diversity metric) at the *landscape* level. Shannon's evenness index is calculated as the ratio between the actual Shannon's diversity index and and the theoretical maximum of the Shannon diversity index ($SHEI = 0$ when only one patch is present and $SHEI = 1$ when the proportion of classes is completely equally distributed).

* `lsm_l_pd` is from the `landscapemetrics` package and can be used to calculate the patch density (an aggregation metric) at the *landscape* level. Patch density describes the fragmentation the landscape and is calculated as the number of patches ($N$) divided by the total landscape area ($A$ [e.g. in $\textrm{m}^2$]).

```{r, eval=FALSE, include=FALSE}
# patch-based metrics are calculated at the patch level (individual patches)
# class-based metrics are calculated at the class level (individual class)
# landscape scale combines all of the classes (ignores cover type)
list_lsm() %>% kableExtra::kable(format = "html") %>% kableExtra::kable_styling(font_size = 11)

landcover <- levels(nlcd)[[1]]

hey_landcovers <- c(
  "open water"
  , "developed, open space"
  , "developed, low intensity"
  , "developed, high intensity"
  , "barren land"
  , "deciduous forest"
  , "evergreen forest"
  , "mixed forest"
  , "shrub/scrub"
  , "grassland herbaceous"
  , "pasture/hay"
  , "cultivated crops"
  , "woody wetlands"
  , "emergent herbaceous wetlands"
) %>% stringr::str_to_title()

landcover[,"landcover"] <- hey_landcovers

per_landcover_type$class_name <- hey_landcovers

per_landcover_type$class_name <- factor(per_landcover_type$class_name, levels = hey_landcovers)


```

---

# Part 1

## Question 2 

What is the grain and extent of the Fort Collins raster? (2 pts)

```{r, results='hide'}
# load data
nlcd <- terra::rast("../data/fort_collins.tif")
# check it
terra::crs(nlcd)
terra::xres(nlcd)
terra::yres(nlcd)
terra::ext(nlcd)
terra::expanse(nlcd) / 10000 # ha
# x
x_temp <- (terra::ext(nlcd)[2] - terra::ext(nlcd)[1]) / 1000
  # ncol(nlcd) * terra::xres(nlcd) / 1000
# y
y_temp <- (terra::ext(nlcd)[4] - terra::ext(nlcd)[3]) / 1000
  # nrow(nlcd) * terra::yres(nlcd) / 1000
```

<span style="color: teal;">
The NLCD raster of Fort Collins has a grain size of **`r paste0(terra::xres(nlcd), "m x ", terra::yres(nlcd), "m")`** with an extent of **`r paste0(round(x_temp,1), "km x ", round(y_temp, 1), "km")`** and covers an expanse of **`r (terra::expanse(nlcd) / 10000) %>% scales::comma(accuracy = 0.1)` ha**.
</span>

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Question 3 

Generate a bar plot showing the proportion of landcover types in the Fort Collins area. Which is most prevalent? Which is least prevalent? Use ggplot for the plotting. (6.5 pts)

### NLCD classification functions

```{r}
# class factor function to return ordered factor from class integer
nlcd_class_fact_fn <- function(class_int){
  return(ordered(class_int
    , levels = FedData::pal_nlcd()$ID
    , labels = FedData::pal_nlcd()$Class
    )
  )
} 
# type factor function to return ordered factor from class integer
nlcd_type_fact_fn <- function(class_int){
  return(ordered(class_int
    , levels = c(11,12,21,22,23,24,31,41,42,43,51,52,71,72,73,74,81,82,90,95)
    , labels = c(
      "Water"
      , "Water"
      , "Developed"
      , "Developed"
      , "Developed"
      , "Developed"
      , "Barren"
      , "Forest"
      , "Forest"
      , "Forest"
      , "Shrubland"
      , "Shrubland"
      , "Herbaceous"
      , "Herbaceous"
      , "Herbaceous"
      , "Herbaceous"
      , "Planted/Cultivated"
      , "Planted/Cultivated"
      , "Wetlands"
      , "Wetlands"
      )
    )
  )
}
# create color pallette based on classes in data
fc_color_pal <- FedData::pal_nlcd() %>% 
  dplyr::filter(ID %in% unique(nlcd)[[1]]) %>% 
  dplyr::pull(Color)
```

### plot

```{r}
# aggregate percent land by class and apply functions
landscapemetrics::lsm_c_pland(nlcd) %>% 
  dplyr::mutate(
    nlcd_class = nlcd_class_fact_fn(class)
    , nlcd_type = nlcd_type_fact_fn(class)
  ) %>%
  ggplot(.) +
    geom_col(mapping = aes(x = value, y = reorder(nlcd_class, -class), fill = nlcd_class), width = 0.7) +
    scale_fill_manual(values = fc_color_pal) +
    geom_text(
      mapping = aes(
        x = value
        , y = reorder(nlcd_class, -class)
        , label = scales::percent(value, scale = 1, accuracy = .1)
      )
      , color = "black", size = 2.5
      , vjust = -0.2
      , hjust = -0.1
    ) +
    scale_x_continuous(expand = expansion(mult = c(0, .1)), labels = scales::percent_format(scale = 1)) +
    labs(
        x = "Percent Land (%)"
        , y = "NLCD Class"
    ) +
    theme_bw() +
    theme(
      legend.position = "none"
    )

```

## Question 4

Calculate the number of evergreen forest patches in the greater Fort Collins region using: 4-neighborhood rule and the 8-neighborhood rule. (2.5 pts) 

```{r}
# calculate patch metrics then summarize to landscape
# or just do class based metric to get it
# lsm_c_np
```

## Question 5

What is the mean pasture/hay patch size in hectare and km2 (using the 4-neighborhood rule): (2.5 pts) 

```{r}
# native units is ha
# lsm_c_area_mn()
```

## Question 6

Of the open water patch-based metrics (using the 4-neighborhood rule), which have the strongest negative correlations? Include the top-three pairings. (3 pts) 

```{r}
# list the top 3 most negative
# write out what these are so that we know what the strongest is
# lsm_p_frac()
```


## Question 7

Generate a plot showing the relationship between the number of open water patches (using the 4-neighborhood rule) and grain size. Use 30m, 60m, 120m, 240, and 600m grain resolution. Use the modal rule and generate your plot using ggplot. (10 pts)

```{r}
# use different aggregations
```

---

# Part 2

For questions 8-10, reclassify “fort_collins.tif” to the scheme outlined on lines 44 to 51. Calculate the following metrics: 

## Question 8

Number of agriculture patches in the greater Fort Collins region using: 4-neighborhood rule & 8-neighborhood rule: (4.5 pts) 

```{r}
# see practice code
# condense 15 cover types down to 6 cover types
```


## Question 9

Calculate the mean Euclidean nearest neighbor distance of agriculture patches (Using the 4-neighborhood rule): (3 pts)

```{r}

```

## Question 10

Include a plot (using the landscape metrics package, e.g., show_lsm) of developed and open water landcover types showing the area in hectares (Using the 4-neighborhood rule). (2 pts)

```{r}

```

---

# Part 3

Calculate the following landscape metrics (for the original fort_collins.tif, i.e., before reclassification). In addition to quantifying the values, briefly describe the meaning of the values. For instance, what does a “high” value of shannon evenness indicate? 

## Question 11

Land-cover richness: (2 pts)

```{r}
# landscape level
```

## Question 12

Shannon diversity:(3.5 pts)

```{r}
# landscape level
```

## Question 13

Shannon evenness: (3.5 pts)

```{r}
# landscape level
```

