---
title: "Lab 4"
subtitle: "ECOL 620 - Applications in Landscape Ecology"
author: "George Woolsey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    # code_folding: hide
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding){ 
    out_dir <- '../';
    rmarkdown::render(inputFile, encoding = encoding, output_file=file.path(dirname(inputFile), out_dir, 'index.html'));
    file.copy(from = "..//index.html", to = '..///data//lab4_george_woolsey.html', overwrite = TRUE)
  })
  
---

# Setup

```{r, include=FALSE, warning=F, message=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results='hide'
  , fig.width = 10
  , fig.height = 7
)
```

```{r, eval=T}
# bread-and-butter
library(tidyverse)
library(lubridate)
library(viridis)
library(scales)
library(latex2exp)
# visualization
library(ggrepel)
library(cowplot)
library(kableExtra)
# spatial analysis
library(USAboundaries)
library(sf)
library(terra)
library(landscapemetrics) #for patch, class, and landscape metrics
# set seed
set.seed(11)
```

# Rules

1. Using the provided R Markdown template, answer all questions and show your R code where necessary. Note, some questions are just plain text written answers.
2. Complete your assignment using the R markdown file and submit individual assignments to Canvas. Knit your script and submit an .html file on Canvas. Please use  the following naming convention: lab4_firstname_lastname.html (ex. lab4_kyle_horton.html).

# Short Answer

## Question 1

Briefly describe (1-2 sentences) what each of the following functions achieve. All of these functions are seen throughout the Lab #3 example code. (1 pt each). 

* `calculate_lsm` is from the `landscapemetrics` package and can be used to calculate several landscape metrics at the user-defined level (i.e. patch, class, or landscape). Landscape metrics include patch area, number of patches, percentage of landscape, and radius of gyration, for example (see `landscapemetrics::list_lsm`).

* `lsm_c_np` is from the `landscapemetrics` package and can be used to calculate the total number of patches (an aggregate metric) at the *class* level.

* `show_cores` is from the `landscapemetrics` package and can be used to create a plot with the core area of patches where the core area is the area of the patch excluding edges of user-defined depth.

* `lsm_l_shei` is from the `landscapemetrics` package and can be used to calculate Shannon's evenness index (a diversity metric) at the *landscape* level. Shannon's evenness index is calculated as the ratio between the actual Shannon's diversity index and and the theoretical maximum of the Shannon diversity index ($SHEI = 0$ when only one patch is present and $SHEI = 1$ when the proportion of classes is completely equally distributed).

* `lsm_l_pd` is from the `landscapemetrics` package and can be used to calculate the patch density (an aggregation metric) at the *landscape* level. Patch density describes the fragmentation the landscape and is calculated as the number of patches ($N$) divided by the total landscape area ($A$ [e.g. in $\textrm{m}^2$]).

```{r, eval=FALSE, include=FALSE}
# patch-based metrics are calculated at the patch level (individual patches)
# class-based metrics are calculated at the class level (individual class)
# landscape scale combines all of the classes (ignores cover type)
list_lsm() %>% kableExtra::kable(format = "html") %>% kableExtra::kable_styling(font_size = 11)
```

---

# Part 1

## Question 2 

What is the grain and extent of the Fort Collins raster? (2 pts)

```{r}

```

## Question 3 

Generate a bar plot showing the proportion of landcover types in the Fort Collins area. Which is most prevalent? Which is least prevalent? Use ggplot for the plotting. (6.5 pts)

```{r}

```

## Question 4

Calculate the number of evergreen forest patches in the greater Fort Collins region using: 4-neighborhood rule and the 8-neighborhood rule. (2.5 pts) 

```{r}

```

## Question 5

What is the mean pasture/hay patch size in hectare and km2 (using the 4-neighborhood rule): (2.5 pts) 

```{r}

```

## Question 6

Of the open water patch-based metrics (using the 4-neighborhood rule), which have the strongest negative correlations? Include the top-three pairings. (3 pts) 

```{r}

```


## Question 7

Generate a plot showing the relationship between the number of open water patches (using the 4-neighborhood rule) and grain size. Use 30m, 60m, 120m, 240, and 600m grain resolution. Use the modal rule and generate your plot using ggplot. (10 pts)

```{r}

```

---

# Part 2

For questions 8-10, reclassify “fort_collins.tif” to the scheme outlined on lines 44 to 51. Calculate the following metrics: 

## Question 8

Number of agriculture patches in the greater Fort Collins region using: 4-neighborhood rule & 8-neighborhood rule: (4.5 pts) 

```{r}

```


## Question 9

Calculate the mean Euclidean nearest neighbor distance of agriculture patches (Using the 4-neighborhood rule): (3 pts)

```{r}

```

## Question 10

Include a plot (using the landscape metrics package, e.g., show_lsm) of developed and open water landcover types showing the area in hectares (Using the 4-neighborhood rule). (2 pts)

```{r}

```

---

# Part 3

Calculate the following landscape metrics (for the original fort_collins.tif, i.e., before reclassification). In addition to quantifying the values, briefly describe the meaning of the values. For instance, what does a “high” value of shannon evenness indicate? 

## Question 11

Land-cover richness: (2 pts)

```{r}

```

## Question 12

Shannon diversity:(3.5 pts)

```{r}

```

## Question 13

Shannon evenness: (3.5 pts)

```{r}

```

