---
title: "Lab 4"
subtitle: "ECOL 620 - Applications in Landscape Ecology"
author: "George Woolsey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    # code_folding: hide
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding){ 
    out_dir <- '../';
    rmarkdown::render(inputFile, encoding = encoding, output_file=file.path(dirname(inputFile), out_dir, 'index.html'));
    file.copy(from = "..//index.html", to = '..///data//lab4_george_woolsey.html', overwrite = TRUE)
  })
  
---

# Setup

```{r, include=FALSE, warning=F, message=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results='hide'
  , fig.width = 10
  , fig.height = 7
)
```

```{r, eval=T}
# bread-and-butter
library(tidyverse)
library(lubridate)
library(viridis)
library(scales)
library(latex2exp)
# visualization
library(ggrepel)
library(cowplot)
library(kableExtra)
# spatial analysis
library(USAboundaries)
library(sf)
library(terra)
library(landscapemetrics) #for patch, class, and landscape metrics
library(FedData) #for nlcd color pallette
# set seed
set.seed(11)
```

# Rules

1. Using the provided R Markdown template, answer all questions and show your R code where necessary. Note, some questions are just plain text written answers.
2. Complete your assignment using the R markdown file and submit individual assignments to Canvas. Knit your script and submit an .html file on Canvas. Please use  the following naming convention: lab4_firstname_lastname.html (ex. lab4_kyle_horton.html).

# Short Answer

## Question 1

Briefly describe (1-2 sentences) what each of the following functions achieve. All of these functions are seen throughout the Lab #3 example code. (1 pt each). 

* `calculate_lsm` is from the `landscapemetrics` package and can be used to calculate several landscape metrics at the user-defined level (i.e. patch, class, or landscape). Landscape metrics include patch area, number of patches, percentage of landscape, and radius of gyration, for example (see `landscapemetrics::list_lsm`).

* `lsm_c_np` is from the `landscapemetrics` package and can be used to calculate the total number of patches (an aggregate metric) at the *class* level.

* `show_cores` is from the `landscapemetrics` package and can be used to create a plot with the core area of patches where the core area is the area of the patch excluding edges of user-defined depth.

* `lsm_l_shei` is from the `landscapemetrics` package and can be used to calculate Shannon's evenness index (a diversity metric) at the *landscape* level. Shannon's evenness index is calculated as the ratio between the actual Shannon's diversity index and and the theoretical maximum of the Shannon diversity index ($SHEI = 0$ when only one patch is present and $SHEI = 1$ when the proportion of classes is completely equally distributed).

* `lsm_l_pd` is from the `landscapemetrics` package and can be used to calculate the patch density (an aggregation metric) at the *landscape* level. Patch density describes the fragmentation the landscape and is calculated as the number of patches ($N$) divided by the total landscape area ($A$ [e.g. in m^2^]).

```{r, eval=FALSE, include=FALSE}
# patch-based metrics are calculated at the patch level (individual patches)
# class-based metrics are calculated at the class level (individual class)
# landscape scale combines all of the classes (ignores cover type)
list_lsm() %>% kableExtra::kable(format = "html") %>% kableExtra::kable_styling(font_size = 11)

landcover <- levels(nlcd)[[1]]

hey_landcovers <- c(
  "open water"
  , "developed, open space"
  , "developed, low intensity"
  , "developed, high intensity"
  , "barren land"
  , "deciduous forest"
  , "evergreen forest"
  , "mixed forest"
  , "shrub/scrub"
  , "grassland herbaceous"
  , "pasture/hay"
  , "cultivated crops"
  , "woody wetlands"
  , "emergent herbaceous wetlands"
) %>% stringr::str_to_title()

landcover[,"landcover"] <- hey_landcovers

per_landcover_type$class_name <- hey_landcovers

per_landcover_type$class_name <- factor(per_landcover_type$class_name, levels = hey_landcovers)


```

---

# Part 1

## Question 2 

What is the grain and extent of the Fort Collins raster? (2 pts)

```{r, results='hide'}
# load data
nlcd <- terra::rast("../data/fort_collins.tif")
# check it
terra::crs(nlcd)
terra::xres(nlcd)
terra::yres(nlcd)
terra::ext(nlcd)
# terra::expanse(nlcd) / 10000 # ha
# x
x_temp <- (terra::ext(nlcd)[2] - terra::ext(nlcd)[1]) / 1000
  # ncol(nlcd) * terra::xres(nlcd) / 1000
# y
y_temp <- (terra::ext(nlcd)[4] - terra::ext(nlcd)[3]) / 1000
  # nrow(nlcd) * terra::yres(nlcd) / 1000
```

<span style="color: teal;">
The NLCD raster of Fort Collins has a grain size of **`r paste0(terra::xres(nlcd), "m x ", terra::yres(nlcd), "m")`** with an extent of **`r paste0(round(x_temp,1), "km x ", round(y_temp, 1), "km")`** and covers an expanse of **`r (terra::expanse(nlcd) / 10000) %>% scales::comma(accuracy = 0.1)` ha**.
</span>

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### NLCD classification functions

```{r}
# class factor function to return ordered factor from class integer
nlcd_class_fact_fn <- function(class_int){
  return(ordered(class_int
    , levels = FedData::pal_nlcd()$ID
    , labels = FedData::pal_nlcd()$Class
    )
  )
} 
# type factor function to return ordered factor from class integer
nlcd_type_fact_fn <- function(class_int){
  return(ordered(class_int
    , levels = c(11,12,21,22,23,24,31,41,42,43,51,52,71,72,73,74,81,82,90,95)
    , labels = c(
      "Water"
      , "Water"
      , "Developed"
      , "Developed"
      , "Developed"
      , "Developed"
      , "Barren"
      , "Forest"
      , "Forest"
      , "Forest"
      , "Shrubland"
      , "Shrubland"
      , "Herbaceous"
      , "Herbaceous"
      , "Herbaceous"
      , "Herbaceous"
      , "Planted/Cultivated"
      , "Planted/Cultivated"
      , "Wetlands"
      , "Wetlands"
      )
    )
  )
}
# FOR LAB type factor function to return ordered factor from class integer
nlcd_labtype_fact_fn <- function(class_int){
  return(ordered(class_int
    , levels = c(41,42,43,21,22,23,24,81,82,71,72,73,74,31,51,52,90,95,11,12)
    , labels = c(
      "Forest"
      , "Forest"
      , "Forest"
      , "Developed"
      , "Developed"
      , "Developed"
      , "Developed"
      , "Agriculture"
      , "Agriculture"
      , "Herbaceous"
      , "Herbaceous"
      , "Herbaceous"
      , "Herbaceous"
      , "Open"
      , "Open"
      , "Open"
      , "Wetlands"
      , "Wetlands"
      , "Water"
      , "Water"
      )
    )
  )
}
# create color pallette based on classes in data
fc_color_pal <- FedData::pal_nlcd() %>% 
  dplyr::filter(ID %in% unique(nlcd)[[1]]) %>% 
  dplyr::pull(Color)
fc_color_type_pal <-
  FedData::pal_nlcd() %>% 
  dplyr::filter(ID %in% unique(nlcd)[[1]]) %>% 
  dplyr::mutate(type = nlcd_type_fact_fn(ID)) %>% 
  dplyr::group_by(type) %>% 
  dplyr::filter(
    n() == 1
    | (n() > 1 & dplyr::row_number() == 2)
  ) %>% 
  dplyr::arrange(type) %>% 
  dplyr::pull(Color)
fc_color_labtype_pal <-
  FedData::pal_nlcd() %>% 
  dplyr::filter(ID %in% unique(nlcd)[[1]]) %>% 
  dplyr::mutate(type = nlcd_labtype_fact_fn(ID)) %>% 
  dplyr::group_by(type) %>% 
  dplyr::filter(
    n() == 1
    | (n() > 1 & dplyr::row_number() == 2)
  ) %>% 
  dplyr::arrange(type) %>% 
  dplyr::pull(Color)
```

## Question 3 

Generate a bar plot showing the proportion of landcover types in the Fort Collins area. Which is most prevalent? Which is least prevalent? Use ggplot for the plotting. (6.5 pts)

```{r}
# aggregate percent land by class and apply functions
nlcd_c_pland_temp <- landscapemetrics::lsm_c_pland(nlcd) %>% 
  dplyr::mutate(
    nlcd_class = nlcd_class_fact_fn(class)
    , nlcd_type = nlcd_type_fact_fn(class)
  )
# plot
  ggplot(nlcd_c_pland_temp) +
    geom_col(mapping = aes(x = value, y = reorder(nlcd_class, -class), fill = nlcd_class), width = 0.7) +
    scale_fill_manual(values = fc_color_pal) +
    geom_text(
      mapping = aes(
        x = value
        , y = reorder(nlcd_class, -class)
        , label = scales::percent(value, scale = 1, accuracy = .1)
      )
      , color = "black", size = 3.5
      , hjust = -0.1
    ) +
    scale_x_continuous(expand = expansion(mult = c(0, .1)), labels = scales::percent_format(scale = 1)) +
    labs(
        x = "Percent Land (%)"
        , y = "NLCD Class"
    ) +
    theme_bw() +
    theme(
      legend.position = "none"
      , axis.title = element_text(size = 8)
      , axis.text.x = element_text(size = 9)
      , panel.grid.major.y = element_blank()
    )

```

<span style="color: teal;">
Based on the NLCD raster of Fort Collins, the landcover class that covers the largest area is **`r nlcd_c_pland_temp %>%  dplyr::slice_max(order_by = value, n = 1) %>% dplyr::pull(nlcd_class) %>% as.character()`** which covers **`r nlcd_c_pland_temp %>%  dplyr::slice_max(order_by = value, n = 1) %>% dplyr::pull(value) %>% scales::percent(scale = 1)`** of the raster extent. The landcover class that covers the smallest area is **`r nlcd_c_pland_temp %>%  dplyr::slice_min(order_by = value, n = 1) %>% dplyr::pull(nlcd_class) %>% as.character()`** which covers **`r nlcd_c_pland_temp %>%  dplyr::slice_min(order_by = value, n = 1) %>% dplyr::pull(value) %>% scales::percent(scale = 1, accuracy = 0.01)`** of the raster extent.
</span>

## Question 4

Calculate the number of evergreen forest patches in the greater Fort Collins region using: 4-neighborhood rule and the 8-neighborhood rule. (2.5 pts) 

```{r}
get_np_fn <- function(my_dir_rule, my_class_str) {
  return(
    landscapemetrics::lsm_c_np(nlcd, directions = my_dir_rule) %>% 
    dplyr::mutate(
      nlcd_class = nlcd_class_fact_fn(class)
      , nlcd_type = nlcd_type_fact_fn(class)
    ) %>% 
    dplyr::filter(
      as.character(nlcd_class) %>% toupper() %in% toupper(my_class_str)
    ) %>% 
    dplyr::mutate(
      patch_direction_rule = paste0(my_dir_rule,"-neighborhood rule")
    ) 
  )
}
# create table for answer
c(4,8) %>% 
  purrr::map_dfr(get_np_fn, my_class_str = c("evergreen forest")) %>% 
  dplyr::select(nlcd_class, patch_direction_rule, value) %>% 
  dplyr::arrange(nlcd_class, patch_direction_rule) %>% 
  dplyr::mutate(value = value %>% scales::comma(accuracy = 1)) %>% 
  kableExtra::kable(
      format = "html" 
      , caption = "Number of patches using different patch rules"
      , escape = FALSE
      , digits = 2
      , col.names = c(
        "NLCD Class"
        , "Patch Rule"
        , "# Patches"
      )
    ) %>% 
    kable_styling(font_size = 11)

```

<span style="color: teal;">
The table above details the number of patches of evergreen forest using a 4-neighborhood rule and a 8-neighborhood rule based on the NLCD raster of Fort Collins.
</span>


## Question 5

What is the mean pasture/hay patch size in hectare and km2 (using the 4-neighborhood rule): (2.5 pts) 

```{r}
# native units is ha
areamn_temp <- landscapemetrics::lsm_c_area_mn(nlcd, directions = 4) %>% 
  dplyr::mutate(
    nlcd_class = nlcd_class_fact_fn(class)
    , nlcd_type = nlcd_type_fact_fn(class)
  ) %>% 
  dplyr::filter(
    as.character(nlcd_class) %>% toupper() %in% toupper("pasture/hay")
  ) %>% 
  dplyr::pull(value)
```

<span style="color: teal;">
The mean patch size of pasture/hay patches in the NLCD raster of Fort Collins is **`r areamn_temp %>% round(1)`** ha (**`r (areamn_temp / 100) %>% round(3)`** km^2^).
</span>

## Question 6

Of the open water patch-based metrics (using the 4-neighborhood rule), which have the strongest negative correlations? Include the top-three pairings. (3 pts) 

```{r, results='asis'}
# get landscape metrics in data frame
water_lsm_temp <- landscapemetrics::calculate_lsm(nlcd, level = "patch", directions = 4, full_name = TRUE) %>% 
  dplyr::mutate(
    nlcd_class = nlcd_class_fact_fn(class)
    , nlcd_type = nlcd_type_fact_fn(class)
  ) %>% 
  dplyr::filter(
    as.character(nlcd_class) %>% toupper() %in% toupper("open water")
  ) %>% 
  dplyr::select(id, function_name, value, nlcd_class) %>% 
  # pivot wider
  tidyr::pivot_wider(
    names_from = function_name
    , values_from = value
  )
# calculate correlation matrix
  cor_matrix <- water_lsm_temp %>% 
    dplyr::select(-c(id, nlcd_class)) %>% 
    cor()
  # keep lower triangle
  cor_matrix[!lower.tri(cor_matrix, diag = FALSE)] <- NA
  cor_matrix <- cor_matrix[2:nrow(cor_matrix), 1:ncol(cor_matrix)-1]
  # create data frame of correlation matrix
  cor_matrix_df <- cor_matrix %>% 
    as.data.frame() %>% 
    tibble::rownames_to_column(var = "var1") %>% 
    tidyr::pivot_longer(
      cols = c(-var1)
      , names_to = "var2"
    ) %>% 
    dplyr::filter(!is.na(value)) %>% 
    dplyr::arrange(value) %>% 
    # attach patch metric variable descriptions
    dplyr::left_join(
      landscapemetrics::list_lsm() %>% dplyr::select(name, function_name) %>% rename(name1 = name)
      , by = c("var1" = "function_name")
    ) %>% 
    dplyr::left_join(
      landscapemetrics::list_lsm() %>% dplyr::select(name, function_name) %>% rename(name2 = name)
      , by = c("var2" = "function_name")
    )
  # create table for answer
  cor_matrix_df %>% 
    dplyr::slice_min(order_by = value, n = 3) %>% 
    dplyr::select(name1, name2, value) %>% 
    kableExtra::kable(
      format = "html" 
      , caption = "Top 3 negative correlations in patch-based metrics<br>NLCD class: open water"
      , escape = FALSE
      , digits = 2
      , col.names = c(
        "Metric 1"
        , "Metric 2"
        , "Pearson Corr. Coeff."
      )
    ) %>% 
    kable_styling(font_size = 11)
```

<span style="color: teal;">
The table above includes the top three open water patch-based metrics (using a 4-neighborhood rule) which have the strongest negative correlation.
</span>

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Question 7

Generate a plot showing the relationship between the number of open water patches (using the 4-neighborhood rule) and grain size. Use 30m, 60m, 120m, 240, and 600m grain resolution. Use the modal rule and generate your plot using ggplot. (10 pts)

```{r}
# define function to return data on patches by class, grain, direction
fn_agg_rst <- function(my_res, my_data, my_dir_rule) {
  return(
  if(my_res>terra::xres(my_data)[1]){
    # aggregate raster
      terra::aggregate(my_data
        , fact = (my_res/terra::xres(my_data)[1]) 
        , fun = "modal"
      ) %>% 
    # calculate number of patches
      landscapemetrics::lsm_c_np(directions = my_dir_rule) %>% 
    # add identifying vars
      dplyr::mutate(
        rast_resolution = my_res
        , patch_direction_rule = my_dir_rule
      )
  }else{if(my_res==terra::xres(my_data)[1]){ # no need to agg. if at native res
      my_data %>% 
    # calculate number of patches
      landscapemetrics::lsm_c_np(directions = my_dir_rule) %>% 
    # add identifying vars
      dplyr::mutate(
        rast_resolution = my_res
        , patch_direction_rule = my_dir_rule
      )
  }}
  )
}
# create data frame of spatial variance vs grain using function defined above
df_grain_patch <- c(30, 60, 120, 240, 600)  %>% # define sequence for factors
  # map = apply a function to each element of a list
  purrr::map_dfr(fn_agg_rst, my_data = nlcd, my_dir_rule = 4) %>% 
  # create class factors
  dplyr::mutate(
    nlcd_class = nlcd_class_fact_fn(class)
    , nlcd_type = nlcd_type_fact_fn(class)
  )
# plot
df_grain_patch %>% 
  dplyr::filter(
    nlcd_class %>% toupper() == toupper("open water")
  ) %>% 
ggplot(.) +
  geom_col(
    mapping = aes(x = as.factor(rast_resolution), y = value)
    , width = 0.7
    , fill = FedData::pal_nlcd() %>% 
        dplyr::filter(Class %>% toupper() %in% toupper("open water")) %>% 
        dplyr::pull(Color)
  ) +
  geom_text(
    mapping = aes(x = as.factor(rast_resolution), y = value, label = value)
    , color = "black", size = 3.5
    , vjust = -0.1
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, .1))) +
  labs(
    y = "# open water patches"
    , x = "raster grain size (m)"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
    , axis.title = element_text(size = 9)
    , panel.grid.major.x = element_blank()
  )
```

---

# Part 2

For questions 8-10, reclassify “fort_collins.tif” to the scheme outlined on lines 44 to 51. Calculate the following metrics: 

## Question 8

Number of agriculture patches in the greater Fort Collins region using: 4-neighborhood rule & 8-neighborhood rule: (4.5 pts) 

```{r}
# see practice code
# condense 15 cover types down to 6 cover types
# get nlcd classes
nlcd_classes <- levels(as.factor(nlcd))[[1]]
nlcd_types <- nlcd_labtype_fact_fn(nlcd_classes %>% as.integer())
#create reclassify matrix: first col: original; second: change to
nlcd_reclass_matrix <- cbind(nlcd_classes, nlcd_types)
#reclassify with classify function 
nlcd_rast_type <- terra::classify(as.factor(nlcd), nlcd_reclass_matrix) 
# aggregate patches
get_np_fn <- function(my_dir_rule, my_type_str) {
  return(
    nlcd_rast_type %>%
    landscapemetrics::lsm_c_np(directions = my_dir_rule) %>% 
    dplyr::mutate(
      nlcd_type = ordered(
        class 
        , levels = c(1:length(nlcd_types %>% unique))
        , labels = nlcd_types %>% unique %>% sort
      )
    ) %>% 
    dplyr::filter(
      as.character(nlcd_type) %>% toupper() %in% toupper(my_type_str)
    ) %>% 
    dplyr::mutate(
      patch_direction_rule = paste0(my_dir_rule,"-neighborhood rule")
    ) 
  )
}
# create table for answer
c(4,8) %>% 
  purrr::map_dfr(get_np_fn, my_type_str = c("agriculture")) %>% 
  dplyr::select(nlcd_type, patch_direction_rule, value) %>% 
  dplyr::arrange(nlcd_type, patch_direction_rule) %>% 
  dplyr::mutate(value = value %>% scales::comma(accuracy = 1)) %>% 
  kableExtra::kable(
      format = "html" 
      , caption = "Number of patches using different patch rules"
      , escape = FALSE
      , digits = 2
      , col.names = c(
        "Landcover Type"
        , "Patch Rule"
        , "# Patches"
      )
    ) %>% 
    kable_styling(font_size = 11)
```

<span style="color: teal;">
The table above details the number of agricultural patches using a 4-neighborhood rule and a 8-neighborhood rule based on the NLCD raster of Fort Collins and a grouping of the NLCD classes into landcover type.
</span>

```{r, include=FALSE, eval=FALSE}
nlcd %>% as.data.frame(xy = TRUE) %>% 
  dplyr::rename(class=3) %>% 
  dplyr::mutate(
    nlcd_class = nlcd_class_fact_fn(class)
    , nlcd_type = nlcd_type_fact_fn(class)
  ) %>% 
ggplot(data=.) +
    geom_raster(mapping=aes(x=x, y=y, fill=nlcd_class)) + 
    scale_fill_manual(values = fc_color_pal) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    # labs(title="Forest cover in 500m buffer of example survey site", fill = "is forest", x="", y="") +
    theme_bw() +
    theme(
      legend.position = "top"
      , legend.direction = "horizontal"
      , legend.title = element_blank()
      , legend.text = element_text(size = 6.5)
      , legend.key.size = unit(0.5, "cm")
      , legend.margin = margin(0,0,-8,0)
      , axis.title = element_blank()
      , axis.text = element_text(size = 6)
    )
####
nlcd_rast_type %>% as.data.frame(xy = TRUE) %>% 
  dplyr::rename(class=3) %>% 
  dplyr::mutate(
    nlcd_type = ordered(
      class 
      , levels = c(1:length(nlcd_types %>% unique))
      , labels = nlcd_types %>% unique %>% sort
    )
  ) %>% 
ggplot(data=.) +
    geom_raster(mapping=aes(x=x, y=y, fill=nlcd_type)) + 
    scale_fill_manual(
      values = fc_color_labtype_pal
    ) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    # labs(title="Forest cover in 500m buffer of example survey site", fill = "is forest", x="", y="") +
    theme_bw() +
    theme(
      legend.position = "top"
      , legend.direction = "horizontal"
      , legend.title = element_blank()
      , legend.text = element_text(size = 6.5)
      , legend.key.size = unit(0.5, "cm")
      , legend.margin = margin(0,0,-8,0)
      , axis.title = element_blank()
      , axis.text = element_text(size = 6)
    )
```

## Question 9

Calculate the mean Euclidean nearest neighbor distance of agriculture patches (Using the 4-neighborhood rule): (3 pts)

```{r}

```

## Question 10

Include a plot (using the landscape metrics package, e.g., show_lsm) of developed and open water landcover types showing the area in hectares (Using the 4-neighborhood rule). (2 pts)

```{r}

```

---

# Part 3

Calculate the following landscape metrics (for the original fort_collins.tif, i.e., before reclassification). In addition to quantifying the values, briefly describe the meaning of the values. For instance, what does a “high” value of shannon evenness indicate? 

## Question 11

Land-cover richness: (2 pts)

```{r}
# landscape level
```

## Question 12

Shannon diversity:(3.5 pts)

```{r}
# landscape level
```

## Question 13

Shannon evenness: (3.5 pts)

```{r}
# landscape level
```

